---
title: "Core-PAM — Rastreabilidade Metodológica e Resultados (Living QMD)"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
# Core settings (edit in 00_setup.R; keep a single source of truth)
# source("scripts/00_setup.R")
```

# 1. Study design (Freeze)
## 1.1 Objectives and endpoints
- Primary endpoint: OS / DSS (per cohort manifest)
- Secondary endpoints: documented in `01_docs/registry/cohort_manifest.csv`

## 1.2 Cohorts
```{r}
# Example: read cohort manifest (edit path if needed)
# library(readr)
# cm <- read_csv("01_docs/registry/cohort_manifest.csv", show_col_types = FALSE)
# cm
```

# 2. Data sources and acquisition
Describe, for each cohort:
- source repository
- files downloaded (raw)
- integrity checks (SHA-256)
- transformations into processed files

# 3. Preprocessing
## 3.1 Expression
- RNA-seq: counts -> TMM/logCPM -> Z-score intra-cohort
- Microarray: log2 intensity -> Z-score intra-cohort
- gene mapping rules (HGNC, probes collapsing)
- effective gene count rules

## 3.2 Clinical
- endpoint mapping rules
- time unit conversion (days -> months by 30.4375)
- reverse Kaplan–Meier follow-up computation

# 4. Model derivation (Core-PAM)
```{r}
# Example: load derivation summary
# library(jsonlite)
# deriv <- fromJSON("results/pamxx/selected_CorePAM_summary.json")
# deriv
```

# 5. Validation (OS/DSS)
## 5.1 Per-cohort results
```{r}
# Example: import final tables (replace with actual filenames)
# library(readr)
# tab_surv <- read_csv("results/main/Table3_Survival_Performance_ByCohort.csv", show_col_types = FALSE)
# tab_surv
```

## 5.2 Meta-analysis
```{r}
# Example: meta summary
# meta <- read_csv("results/main/Meta_Survival_Summary.csv", show_col_types = FALSE)
# meta
```

# 6. QC and forensics
- off-diagonal correlations
- METABRIC PCA forensics
- leakage checks
- PH assumption checks

# 7. Figures and tables (embedded)
```{r}
# knitr::include_graphics("figures/main/Fig1_StudyDesign_Freeze.pdf")
```

# 8. Declarations (journal-ready)
- Ethics approval and consent to participate
- Consent for publication
- Availability of data and materials
- Competing interests
- Funding
- Authors' contributions
- Acknowledgements
