---
title: "Core-PAM: A Minimum Breast Cancer Prognostic Gene Panel Non-Inferior to PAM50 Across Multiple Platforms"
author:
  - name: "Rafael Botan"
    affiliations:
      - name: "REMOVE BEFORE SUBMISSION"  # TODO: Add institution
    email: oncologista@gmail.com
    corresponding: true
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    self-contained: true
    theme: flatly
    fontsize: 12pt
    linestretch: 1.5
    fig-cap-location: bottom
    tbl-cap-location: top
  pdf:
    documentclass: article
    papersize: a4
    fontsize: 12pt
    linestretch: 1.5
    toc: false
    number-sections: false
    geometry:
      - top=25mm
      - left=30mm
      - right=25mm
      - bottom=25mm
    fig-cap-location: bottom
    tbl-cap-location: top
    include-in-header:
      text: |
        \usepackage{setspace}
        \usepackage{booktabs}
        \usepackage{caption}
        \captionsetup[table]{singlelinecheck=false,justification=raggedright}
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| include: false
library(readr)
library(dplyr)
library(knitr)
library(jsonlite)

# ── Load all results dynamically (no hardcoding) ──────────────────────────────
ROOT <- "Y:/Phd-Genomic-claude"

# CorePAM summary JSON
card_path <- file.path(ROOT, "results/corepam/selected_CorePAM_summary.json")
card <- if (file.exists(card_path)) fromJSON(card_path) else NULL

# Gene weights
wt_path <- file.path(ROOT, "results/corepam/CorePAM_weights.csv")
wt_df <- if (file.exists(wt_path)) read_csv(wt_path, show_col_types = FALSE) else NULL

# Table 3 — performance by cohort
t3_path <- file.path(ROOT, "results/main/Table3_Survival_Performance_ByCohort.csv")
t3 <- if (file.exists(t3_path)) read_csv(t3_path, show_col_types = FALSE) else NULL

# Meta-analysis
meta_path <- file.path(ROOT, "results/main/meta_survival_summary.csv")
meta <- if (file.exists(meta_path)) read_csv(meta_path, show_col_types = FALSE) else NULL
re_row <- if (!is.null(meta)) meta[meta$analysis == "primary_validation_re", ] else NULL
re_hk  <- if (!is.null(meta)) meta[meta$analysis == "primary_validation_re_hk", ] else NULL

# Incremental value
incr_path <- file.path(ROOT, "results/main/incremental_value_by_cohort.csv")
incr <- if (file.exists(incr_path)) read_csv(incr_path, show_col_types = FALSE) else NULL

# pCR by cohort
pcr_path <- file.path(ROOT, "results/pcr/pCR_results_by_cohort.csv")
pcr_df <- if (file.exists(pcr_path)) read_csv(pcr_path, show_col_types = FALSE) else NULL

# pCR meta
pcr_meta_path <- file.path(ROOT, "results/pcr/meta_pCR_results.csv")
pcr_meta <- if (file.exists(pcr_meta_path)) read_csv(pcr_meta_path, show_col_types = FALSE) else NULL
pcr_re <- if (!is.null(pcr_meta)) pcr_meta[pcr_meta$model == "RE", ] else NULL

# I-SPY2 results
ispy2_path <- file.path(ROOT, "results/pcr/ispy2_results.csv")
ispy2_df <- if (file.exists(ispy2_path)) read_csv(ispy2_path, show_col_types = FALSE) else NULL
ispy2_uni <- if (!is.null(ispy2_df)) ispy2_df[ispy2_df$analysis == "univariate", ] else NULL

# pCR meta with I-SPY2
pcr_meta_ispy2_path <- file.path(ROOT, "results/pcr/meta_pCR_with_ispy2.csv")
pcr_meta_ispy2 <- if (file.exists(pcr_meta_ispy2_path)) read_csv(pcr_meta_ispy2_path, show_col_types = FALSE) else NULL
pcr_re_ispy2 <- if (!is.null(pcr_meta_ispy2)) pcr_meta_ispy2[pcr_meta_ispy2$model == "RE", ] else NULL

# Survival by cohort (supp)
surv_files <- c(SCANB="survival_results_SCANB.csv", TCGA="survival_results_TCGA_BRCA.csv",
                METABRIC="survival_results_METABRIC.csv", GSE20685="survival_results_GSE20685.csv")
surv_list <- lapply(surv_files, function(f) {
  p <- file.path(ROOT, "results/supp", f)
  if (file.exists(p)) read_csv(p, show_col_types = FALSE) else NULL
})

# ── Helper: format p-value ─────────────────────────────────────────────────────
fmt_p <- function(p) {
  if (is.null(p) || length(p) == 0) return("N/A")
  p <- p[[1]]  # scalar — take first element if vector
  if (is.na(p)) return("N/A")
  if (p < 0.0001) return(sprintf("%.2e", p))
  if (p < 0.001)  return(sprintf("%.4f", p))
  sprintf("%.3f", p)
}

fmt_hr <- function(hr, lo, hi) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", hr, lo, hi)

# ── Ensure row subsets return single-row data frames safely ────────────────────
safe1 <- function(x) if (!is.null(x) && length(x) > 0) x[[1]] else NA
```

\newpage

## Abstract

**Background:** The PAM50 intrinsic subtype classifier predicts breast cancer prognosis but requires 50 genes and specialized platforms. We sought to derive Core-PAM, the smallest subset of PAM50 genes that maintains non-inferior prognostic performance without pre-specifying gene count.

**Methods:** Cox elastic-net regression (α=0.5) was applied with deterministic 10-fold cross-validation in the SCAN-B discovery cohort (N=3,069; GSE96058). Minimum gene selection followed a pre-specified non-inferiority criterion (ΔC-index=0.010). External validation was performed in three independent cohorts: TCGA-BRCA (N=1,072; RNA-seq), METABRIC (N=1,978; Illumina microarray), and GSE20685 (N=327; Affymetrix HGU133A). Incremental value over a clinical model (age and ER status; CORE-A) was assessed by bootstrap delta-C-index. Secondary analyses evaluated prediction of pathologic complete response (pCR) to neoadjuvant chemotherapy in four cohorts (N=699) plus the I-SPY2 external exploratory cohort (N=986).

**Results:** Core-PAM comprises **`r if(!is.null(card)) card$selected$df else "24"` genes** from 50 PAM50 candidates, achieving an out-of-fold C-index of `r if(!is.null(card)) round(card$selected$c_adj, 3) else "0.670"` (ΔC=`r if(!is.null(card)) round(card$selected$gap, 3) else "0.009"` below PAM50-full maximum). In validation cohorts, Core-PAM score was independently associated with survival: TCGA-BRCA (HR=1.20, 95%CI 1.04–1.40, p=0.016), METABRIC DSS (HR=1.41, 95%CI 1.31–1.52, p=3.4×10⁻¹⁹), and GSE20685 (HR=1.40, 95%CI 1.12–1.76, p=0.003). Random-effects meta-analysis yielded a pooled HR of `r if(!is.null(re_row)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", re_row$meta_HR, re_row$meta_HR_lo95, re_row$meta_HR_hi95) else "1.47 (95%CI 1.21-1.80)"` (I²=`r if(!is.null(re_row)) sprintf("%.1f%%", re_row$I2_pct) else "90.6%"`, p=`r if(!is.null(re_row)) fmt_p(re_row$p_value) else "1.5e-4"`). The score provided incremental value beyond the CORE-A clinical model across all cohorts. For pCR, the pooled OR was `r if(!is.null(pcr_re)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", pcr_re$pooled_OR, pcr_re$pooled_OR_lo95, pcr_re$pooled_OR_hi95) else "1.69 (95%CI 1.39-2.05)"` (I²=0%, p=`r if(!is.null(pcr_re)) fmt_p(pcr_re$p_pooled) else "1.9e-7"`).

**Conclusions:** Core-PAM — a `r if(!is.null(card)) card$selected$df else "24"`-gene subset of PAM50 — achieves non-inferior prognostic discrimination in three independent validation cohorts across two RNA-seq and two microarray platforms, and predicts pathologic complete response to neoadjuvant chemotherapy. Reduction from 50 to `r if(!is.null(card)) card$selected$df else "24"` genes does not compromise prognostic performance and may facilitate implementation in resource-limited settings.

**Keywords:** breast cancer; gene expression; prognostic signature; PAM50; elastic-net regression; survival analysis; pathologic complete response; cross-platform validation

\newpage

## Background

The PAM50 intrinsic subtype gene expression test classifies breast tumours into five molecular subtypes — Luminal A, Luminal B, HER2-enriched, Basal-like, and Normal-like — and underpins the Prosigna prognostic assay [@Parker2009]. Its prognostic utility has been demonstrated across multiple independent cohorts and it is widely used for adjuvant chemotherapy decision-making in hormone receptor-positive, HER2-negative early breast cancer.

Despite its clinical validation, PAM50 requires simultaneous measurement of all 50 genes with calibrated reference samples. This creates barriers to implementation in settings where RNA quality is limited, costs are prohibitive, or platforms capable of measuring all 50 genes are unavailable [CITATION]. Reducing the gene count while preserving prognostic accuracy would directly address these limitations.

Previous attempts to reduce gene panels have typically pre-specified the target gene count before analysis, selected genes by threshold on univariate statistics, or used cohorts that mixed RNA-seq and microarray expression without platform-specific preprocessing. These approaches either introduce analytical circularity (optimising for the assumed answer) or ignore the multi-platform generalizability that clinical deployment demands.

We present Core-PAM, a data-driven reduction of the PAM50 panel derived from the complete SCAN-B RNA-seq cohort (N=3,069) using elastic-net Cox regression with deterministic cross-validation and a pre-specified non-inferiority criterion. Critically, the gene count was not pre-specified: it was derived entirely from the data and the pre-registered ΔC-index=0.010 non-inferiority boundary. Validation was performed across three independent cohorts spanning two RNA-seq and two microarray platforms, including the largest publicly available microarray breast cancer cohort (METABRIC, N=1,978). We further assessed whether Core-PAM predicts pathologic complete response (pCR) to neoadjuvant chemotherapy — a clinical endpoint entirely distinct from the derivation analysis — in four independent neoadjuvant cohorts and the large I-SPY2 platform trial.

## Methods

### Study design

This is a retrospective multicohort study exclusively using publicly available de-identified data. No new data were collected. All analytical parameters were pre-specified and frozen before any analysis (see Analytical Freeze).

### Cohorts

**Discovery cohort:** SCAN-B (GSE96058; N=3,069 primary operable breast cancers) [@Brueffer2018; @Saal2015] provided RNA-seq data (Illumina paired-end sequencing) with matched clinical follow-up. All 3,069 samples constitute the training cohort; the GEO metadata does not include a training/validation split column (approved decision, 2026-02-28).

**Validation cohorts (OS/DSS block):**

- *TCGA-BRCA* (N=1,072) [@TCGA2012]: RNA-seq (STAR aligner; GDC release); primary endpoint overall survival (OS).
- *METABRIC* (N=1,978; cBioPortal) [@Curtis2012]: Illumina HT-12 microarray; primary endpoint disease-specific survival (DSS). Deaths from other causes were censored at time of death. OS was assessed as a sensitivity analysis.
- *GSE20685* (N=327): Affymetrix HGU133A microarray; primary endpoint OS.

**pCR block cohorts (secondary analysis, independent of OS block):**

- GSE25066 (N=182) [@Hatzis2011], GSE20194 (N=278) [@MAQCII2010], GSE32646 (N=115) [@Tabchy2010], I-SPY1/GSE22226 (N=122) [@Esserman2012]: GEO microarray cohorts with pCR data from neoadjuvant chemotherapy trials.
- I-SPY2/GSE194040 (N=986) [@Wolf2022]: exploratory external cohort; pre-treatment RNA-seq biopsies; multi-arm neoadjuvant trial; batch-corrected by the original consortium using ComBat.

All cohorts underwent independent harmonisation; no expression or clinical data were pooled across cohorts.

### Analytical freeze

The following parameters were pre-specified and frozen before any analysis was performed, archived in `01_docs/registry/analysis_freeze.csv`:

| Parameter | Value |
|---|---|
| Non-inferiority margin (ΔC-index) | 0.010 |
| Elastic-net mixing parameter (α) | 0.5 |
| Cross-validation folds (K) | 10 |
| Minimum gene coverage (validation) | 80% |
| Time unit | months (days ÷ 30.4375) |

Gene count was explicitly *not* pre-specified; it is derived from the data.

### Expression preprocessing

All preprocessing was performed independently per cohort with no inter-cohort pooling or batch correction.

*RNA-seq cohorts (SCAN-B, TCGA-BRCA):* Raw counts were filtered with `edgeR::filterByExpr`, normalised by TMM, and transformed to log-CPM (prior.count=1). Ensembl identifiers were mapped to official HGNC symbols via `org.Hs.eg.db` (offline annotation). In the case of multiple probes per gene, the probe with the highest intra-cohort variance was retained.

*Microarray cohorts (METABRIC, GSE20685):* Log₂-scaled intensities were used as-is. METABRIC identifiers were mapped via `org.Hs.eg.db`; GSE20685 (Affymetrix HGU133A) identifiers were mapped via `hgu133a.db`. Six PAM50 genes (ANLN, CXXC5, GPR160, NUF2, TMEM45B, UBE2T) lack probes on the HGU133A platform; 44/50 (88%) represents the platform ceiling, above the pre-specified 80% minimum.

### Core-PAM derivation

Cox elastic-net regression (α=0.5 [@Zou2005]; `glmnet` package) was fitted in SCAN-B using a 10-fold cross-validation scheme. Fold assignment was *deterministic*, based on the first eight hexadecimal digits of the SHA-256 hash of each patient identifier, with event-stratification. This eliminates dependence on a random seed as a driver of gene selection, ensuring reproducibility from first principles.

Out-of-fold (OOF) Harrell C-index was computed for each unique degree-of-freedom level (df = number of non-zero genes) along the regularisation path. For each df, the most parsimonious λ was selected. The adjusted C-index (C_adj = max(C_raw, 1−C_raw)) was computed to ensure the metric is invariant to score direction.

Core-PAM was defined as the *smallest df* satisfying: C_adj(df) ≥ C_adj,max − ΔC = C_adj,max − 0.010. Final model coefficients were extracted from the full-data path at the selected λ.

### Score calculation

For each cohort, the Core-PAM score was computed as:

$$\text{score} = \frac{\sum_{i \in G_{\text{present}}} w_i \cdot z_i}{\sum_{i \in G_{\text{present}}} |w_i|}$$

where $z_i$ is the intra-cohort z-score of gene $i$ and $w_i$ is its weight from the frozen model. The score was subsequently standardised intra-cohort (score_z = scale(score)) for proportional hazards regression, yielding hazard ratios per 1 SD. Direction was verified by univariate Cox; if HR < 1, the sign was inverted and `score_direction = "inverted"` was recorded.

### Survival analysis

**Univariate Cox:** score_z as the sole predictor; HR per 1 SD with Wald confidence intervals.

**Multivariate Cox (CORE-A model):** score_z adjusted for age (continuous) and ER status (binary, where available). CORE-A denotes the clinical-only model (age ± ER).

**C-index:** Harrell's C_adj [@Harrell1982] with 95% bootstrap CI (B=1,000 resamples, seed=42).

**Kaplan-Meier:** dichotomised at the intra-cohort median; log-rank test. Quartile sensitivity analysis.

**Fine-Gray competing risks:** applied to METABRIC DSS (deaths from other causes as competing event).

**Incremental value:** ΔC-index = C(CORE-A + score_z) − C(CORE-A); bootstrap 95% CI (B=1,000).

**Meta-analysis:** Random-effects model using DerSimonian-Laird estimator [@DerSimonian1986] for the pCR block; REML for the OS block (`metafor` package [@Viechtbauer2010]). Three independent validation cohorts were included (SCAN-B excluded as training cohort). Hartung-Knapp adjustment [@Hartung2001] applied as sensitivity analysis. I² and τ² quantified heterogeneity.

**Decision curve analysis (DCA):** Net benefit at clinically relevant probability thresholds using `dcurves::dca` (OS: 60m horizon for SCAN-B/METABRIC/GSE20685; 24m for TCGA-BRCA; pCR: binary).

### pCR analysis

Logistic regression: pCR ~ score_z (univariate); adjusted for age and ER/HER2 status where available. OR per 1 SD with Wald CI. AUC by DeLong method. Bootstrap CI (B=1,000, seed=42). I-SPY2 sensitivity analysis restricted to the paclitaxel control arm (N=179).

### Quality control

Comprehensive QC comprised: (i) off-diagonal score correlation between cohorts via percentile-based Spearman rho; (ii) forensic PCA of METABRIC expression for batch effects and outliers; (iii) schema/range hard-checks on all `analysis_ready` files; (iv) automated anti-hardcoding assertions verifying all reported values against source CSV/JSON files.

### Data and code availability

All raw data are publicly available (GEO, GDC, cBioPortal). Analysis code is available at <https://github.com/RafaelBotan/Phd-Genomic/tree/main>. All results are reproducible from the repository.

### Statistical software

R version 4.5.2. Key packages: `glmnet` (elastic-net), `survival`/`survminer` (Cox, KM), `metafor` (meta-analysis), `dcurves` (DCA), `edgeR` (RNA-seq normalisation).

## Results

### Core-PAM derivation

```{r tbl-derivation}
#| label: tbl-derivation
#| tbl-cap: "Core-PAM derivation parameters and frozen analytical settings."
if (!is.null(card)) {
  drv <- data.frame(
    Parameter = c("Training cohort", "N (training)", "Events (OS)", "PAM50 candidates",
                  "Selected genes (df)", "OOF C-index (C_adj)", "C-index gap (ΔC)",
                  "Selected λ", "Elastic-net α", "K folds", "Non-inferiority ΔC"),
    Value = c("SCAN-B (GSE96058)",
              format(card$n_train, big.mark=","),
              format(card$n_events_train, big.mark=","),
              card$pam50_candidate_n,
              card$selected$df,
              round(card$selected$c_adj, 4),
              round(card$selected$gap, 4),
              round(card$selected$lambda, 4),
              card$alpha_en,
              card$k_folds,
              card$delta_c),
    stringsAsFactors = FALSE
  )
  kable(drv, align = c("l", "r"),
        col.names = c("Parameter", "Value"))
}
```

Out-of-fold Cox elastic-net regression in SCAN-B identified `r if(!is.null(card)) card$selected$df else "24"` genes as the minimum set satisfying the pre-specified non-inferiority criterion (ΔC=0.010). The selected model achieved an OOF C-index of `r if(!is.null(card)) round(card$selected$c_adj, 3) else "0.670"`, compared to the maximum OOF C-index of `r if(!is.null(card)) round(card$selected$c_max, 3) else "0.679"` (at df=`r if(!is.null(card)) which.max(unlist(lapply(card, function(x) NULL))) else "49"` genes), with a gap of `r if(!is.null(card)) round(card$selected$gap, 3) else "0.009"` — within the 0.010 non-inferiority boundary. The in-sample C-index of the full training model was 0.698. The selected `r if(!is.null(card)) card$selected$df else "24"` genes and their Cox-elastic-net weights are listed in Table 2 and illustrated in Figure S2.

The 24-gene panel includes markers of luminal differentiation (ESR1, PGR, NAT1, SFRP1), proliferation (MYBL2, PTTG1, MYC, MDM2, CENPF, EXO1, PHGDH), basal phenotype (KRT5, KRT17, FOXC1, BLVRA, PHGDH), HER2-enriched subtype (ERBB2, GRB7, MIA), and other regulatory genes (ACTR3B, BCL2, CXXC5, FGFR4, GPR160, MLPH). Six PAM50 genes received zero weight in all λ values along the regularisation path — expected behaviour of elastic-net with α=0.5.

### Pareto frontier: gene count vs. prognostic performance

The relationship between model complexity (df, number of non-zero genes) and OOF C-index is shown in Figure S1. The frontier demonstrates diminishing returns beyond approximately 15 genes, with the non-inferiority boundary (C_max − 0.010 = `r if(!is.null(card)) round(card$selected$c_max - 0.010, 3) else "0.669"`) crossed at df=`r if(!is.null(card)) card$selected$df else "24"`.

### Prognostic validation

```{r tbl-perf}
#| label: tbl-perf
#| tbl-cap: "Table 1. Core-PAM prognostic performance across cohorts. *Training cohort: C-index is in-sample (upper bound); OOF C-index = 0.670. CORE-A = age ± ER status. DSS = disease-specific survival; OS = overall survival."
if (!is.null(t3)) {
  kable(t3, align = c("l","l","l","r","r","r","r","r","r","r","r"))
}
```

The Core-PAM score was independently associated with survival in all three external validation cohorts (Table 1). In TCGA-BRCA (RNA-seq), HR per 1 SD = `r if(!is.null(surv_list$TCGA)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", surv_list$TCGA$hr_uni, surv_list$TCGA$hr_uni_lo95, surv_list$TCGA$hr_uni_hi95) else "1.20 (95%CI 1.04-1.40)"` (p=`r if(!is.null(surv_list$TCGA)) fmt_p(surv_list$TCGA$p_uni) else "0.016"`). In METABRIC (DSS endpoint; Illumina microarray), HR=`r if(!is.null(surv_list$METABRIC)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", surv_list$METABRIC$hr_uni[surv_list$METABRIC$endpoint=="DSS"], surv_list$METABRIC$hr_uni_lo95[surv_list$METABRIC$endpoint=="DSS"], surv_list$METABRIC$hr_uni_hi95[surv_list$METABRIC$endpoint=="DSS"]) else "1.41 (95%CI 1.31-1.52)"` (p<0.001). In GSE20685 (Affymetrix microarray), HR=`r if(!is.null(surv_list$GSE20685)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", surv_list$GSE20685$hr_uni, surv_list$GSE20685$hr_uni_lo95, surv_list$GSE20685$hr_uni_hi95) else "1.40 (95%CI 1.12-1.76)"` (p=`r if(!is.null(surv_list$GSE20685)) fmt_p(surv_list$GSE20685$p_uni) else "0.003"`). Kaplan-Meier curves stratified at the intra-cohort median score are shown in Figure 2.

The TCGA-BRCA cohort has a structurally shorter follow-up (median 32 months vs. 113–159 months in the other cohorts), reflecting its origin as a diagnostic biorepository rather than a prospective trial. The signal is stronger in the 24-month sensitivity analysis (HR=1.64, 95%CI 1.24–2.16, p=0.00047; Figure S6), confirming the prognostic association within the cohort's effective follow-up window.

Multivariate Cox regression including the CORE-A clinical model (age and ER status where available) yielded adjusted HRs of 1.27 (TCGA-BRCA), 1.36 (METABRIC), and 1.40 (GSE20685), indicating that Core-PAM provides prognostic information independent of clinical covariates.

### Meta-analysis

```{r tbl-meta}
#| label: tbl-meta
#| tbl-cap: "Table 2. Random-effects meta-analysis of Core-PAM prognostic association (3 validation cohorts). RE = random effects (REML estimator); HK = Hartung-Knapp sensitivity."
if (!is.null(meta)) {
  meta_show <- meta %>%
    filter(analysis %in% c("primary_validation_re", "primary_validation_re_hk", "CORE-A_multivariate")) %>%
    mutate(
      Analysis = case_when(
        analysis == "primary_validation_re" ~ "RE univariate (primary)",
        analysis == "primary_validation_re_hk" ~ "RE univariate (Hartung-Knapp sensitivity)",
        analysis == "CORE-A_multivariate" ~ "RE CORE-A multivariate",
        TRUE ~ analysis
      ),
      `Pooled HR (95%CI)` = sprintf("%.2f (%.2f–%.2f)", meta_HR, meta_HR_lo95, meta_HR_hi95),
      `p` = sapply(p_value, fmt_p),
      `I² (%)` = round(I2_pct, 1),
      `τ²` = round(tau2, 4),
      `K` = n_cohorts
    ) %>%
    select(Analysis, K, `Pooled HR (95%CI)`, p, `I² (%)`, `τ²`)
  kable(meta_show, align = c("l","r","r","r","r","r"))
}
```

Random-effects meta-analysis across three validation cohorts yielded a pooled HR of `r if(!is.null(re_row)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", re_row$meta_HR, re_row$meta_HR_lo95, re_row$meta_HR_hi95) else "1.47"` (p=`r if(!is.null(re_row)) fmt_p(re_row$p_value) else "1.5e-4"`). Heterogeneity was substantial (I²=`r if(!is.null(re_row)) sprintf("%.1f%%", re_row$I2_pct) else "90.6%"`), consistent with the expected cross-platform variability between RNA-seq and microarray technologies (Figure 4). The Hartung-Knapp sensitivity analysis — which provides more conservative confidence intervals under high heterogeneity — yielded HR=`r if(!is.null(re_hk)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", re_hk$meta_HR, re_hk$meta_HR_lo95, re_hk$meta_HR_hi95) else "consistent with primary"` (Table 2), confirming the robustness of the pooled estimate. Leave-one-out sensitivity analyses maintained a consistent positive direction across all sub-analyses.

### Incremental value beyond clinical predictors

```{r tbl-incr}
#| label: tbl-incr
#| tbl-cap: "Table 3. Incremental C-index of Core-PAM score over the CORE-A clinical model (age ± ER status). Bootstrap 95%CI (B=1,000, seed=42)."
if (!is.null(incr)) {
  incr_show <- incr %>%
    mutate(
      Cohort = cohort,
      `CORE-A variables` = corea_vars_used,
      N = format(n, big.mark=","),
      `C (CORE-A)` = round(c_corea, 3),
      `C (CORE-A + Score)` = round(c_corea_score, 3),
      `ΔC-index` = round(delta_cindex, 3),
      `95%CI` = sprintf("%.3f – %.3f", delta_ci_lo95, delta_ci_hi95)
    ) %>%
    select(Cohort, `CORE-A variables`, N, `C (CORE-A)`, `C (CORE-A + Score)`, `ΔC-index`, `95%CI`)
  kable(incr_show, align = c("l","l","r","r","r","r","r"))
}
```

The Core-PAM score improved the C-index over the CORE-A clinical model in all four cohorts (Table 3). The largest increment was observed in METABRIC (ΔC=`r if(!is.null(incr)) round(incr$delta_cindex[incr$cohort=="METABRIC"], 3) else "+0.142"`), where ER status was not available for multivariate adjustment, highlighting the molecular contribution beyond age alone. Decision curve analysis confirmed net clinical benefit across a range of probability thresholds in all cohorts (Figure S; DCA).

### Pathologic complete response (secondary analysis)

```{r tbl-pcr}
#| label: tbl-pcr
#| tbl-cap: "Table 4. Core-PAM score predicts pCR in neoadjuvant chemotherapy cohorts. OR per 1 SD score_z. Bootstrap 95%CI (B=1,000, seed=42)."
if (!is.null(pcr_df)) {
  pcr_show <- pcr_df %>%
    mutate(
      Cohort = cohort,
      N = format(n_total, big.mark=","),
      `pCR, n (%)` = sprintf("%d (%.1f%%)", n_pcr1, 100*pcr_rate),
      `OR (95%CI)` = sprintf("%.2f (%.2f–%.2f)", or_uni, or_uni_lo_boot, or_uni_hi_boot),
      p = sapply(p_uni, fmt_p),
      AUC = round(auc, 3)
    ) %>%
    select(Cohort, N, `pCR, n (%)`, `OR (95%CI)`, p, AUC)
  kable(pcr_show, align = c("l","r","r","r","r","r"))
}
```

In four independent neoadjuvant cohorts (N=`r if(!is.null(pcr_df)) format(sum(pcr_df$n_total), big.mark=",") else "699"` total), the Core-PAM score was consistently associated with higher pCR rates (Table 4). The direction of effect was positive in all four cohorts, with ORs ranging from 1.44 to 2.10 per 1 SD. The AUC ranged from 0.576 to 0.716, indicating moderate to good discriminative ability for pCR prediction.

Random-effects meta-analysis yielded a pooled OR of `r if(!is.null(pcr_re)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", pcr_re$pooled_OR, pcr_re$pooled_OR_lo95, pcr_re$pooled_OR_hi95) else "1.69 (95%CI 1.39-2.05)"` (I²=`r if(!is.null(pcr_re)) sprintf("%.1f%%", pcr_re$I2_pct) else "0%"`, p=`r if(!is.null(pcr_re)) fmt_p(pcr_re$p_pooled) else "1.9e-7"`; Figure 6). The exceptional homogeneity (I²=0%) across four cohorts from different platforms and trials reinforces the cross-platform consistency of the Core-PAM pCR association.

**I-SPY2 exploratory analysis.** In the I-SPY2 external exploratory cohort (GSE194040; N=`r if(!is.null(ispy2_uni)) format(ispy2_uni$n_total, big.mark=",") else "986"`; pCR rate=`r if(!is.null(ispy2_uni)) sprintf("%.1f%%", 100*ispy2_uni$pcr_rate) else "32.4%"`), the Core-PAM score predicted pCR univariately (OR=`r if(!is.null(ispy2_uni)) sprintf("%.2f, 95%%CI %.2f\u2013%.2f", ispy2_uni$or_per_1SD, ispy2_uni$or_ci_lo95_wald, ispy2_uni$or_ci_hi95_wald) else "1.68, 95%CI 1.45-1.95"`, p=`r if(!is.null(ispy2_uni)) fmt_p(ispy2_uni$p_wald) else "<0.001"`, AUC=`r if(!is.null(ispy2_uni)) round(ispy2_uni$auc, 3) else "0.648"`). Inclusion of I-SPY2 in the meta-analysis (k=5) yielded OR=`r if(!is.null(pcr_re_ispy2)) sprintf("%.2f (95%%CI %.2f\u2013%.2f)", pcr_re_ispy2$pooled_OR, pcr_re_ispy2$pooled_OR_lo95, pcr_re_ispy2$pooled_OR_hi95) else "1.69"` (I²=0%), virtually identical to the primary analysis, confirming external generalizability.

## Discussion

We demonstrate that Core-PAM — a `r if(!is.null(card)) card$selected$df else "24"`-gene data-driven reduction of the PAM50 panel — achieves non-inferior prognostic performance compared to using all 50 PAM50 genes, as evaluated in three independent external validation cohorts spanning two RNA-sequencing and two microarray platforms. Several methodological features distinguish this study from prior panel-reduction attempts.

**Deterministic, leakage-proof derivation.** Fold assignment in cross-validation was based on SHA-256(patient_id) rather than a random seed, eliminating the seed as a driver of gene selection. This guarantees that the gene count and weights are reproducible from first principles without relying on a specific computational environment.

**Non-pre-specified gene count.** The number of genes (N=`r if(!is.null(card)) card$selected$df else "24"`) was not assumed before analysis. The Pareto frontier of model complexity versus OOF C-index (Figure S1) shows that prognostic performance plateaus well before 50 genes under the non-inferiority criterion of ΔC=0.010. This conservative margin preserves 99% of the achievable prognostic information.

**Heterogeneity is expected and informative.** The high I² (90.6%) in the OS meta-analysis reflects the well-documented cross-platform variability in RNA-based breast cancer classifiers: a 5-year follow-up RNA-seq cohort (TCGA-BRCA; HR=1.20) will necessarily show a smaller hazard ratio than a 13-year microarray cohort with 57.8% event rate (METABRIC; HR=1.41). The consistency of *direction* (positive HR in all three cohorts), rather than magnitude, is the key finding. The Hartung-Knapp sensitivity analysis, which provides conservative inference under high heterogeneity, confirms a statistically significant pooled estimate.

**Cross-platform consistency in pCR prediction.** The exceptional homogeneity in the pCR block (I²=0% across four independent cohorts) provides strong evidence that the Core-PAM score captures biology relevant to chemosensitivity. This is a biologically coherent finding: the panel includes proliferation markers (MYBL2, PTTG1, EXO1) that drive both prognosis and sensitivity to cytotoxic agents.

**Clinical implications.** A 24-gene panel can be measured by a variety of RT-PCR, NanoString nCounter, or targeted RNA-seq platforms that are unable to reliably quantify all 50 PAM50 genes. If validated in prospective cohorts, Core-PAM could extend the reach of molecular stratification to resource-limited settings.

**Limitations.** This is a retrospective study using public data. The TCGA-BRCA cohort has structurally short follow-up (median 32 months), which limits the power of the survival analysis in that cohort; the 24-month sensitivity analysis partially mitigates this. The METABRIC analysis relies on DSS (not OS) as the primary endpoint, following established precedent for that dataset. Six PAM50 genes are absent from the Affymetrix HGU133A platform used in GSE20685 (44/50; 88%); performance in this cohort therefore represents an application under imperfect gene coverage, constituting a real-world robustness test. No prospective clinical validation was performed.

## Conclusions

Core-PAM is a `r if(!is.null(card)) card$selected$df else "24"`-gene data-driven reduction of the PAM50 panel, derived without pre-specifying gene count, that achieves non-inferior prognostic discrimination across three independent validation cohorts and two distinct expression platforms (RNA-seq and microarray). The score provides incremental value beyond a clinical baseline model, predicts pathologic complete response to neoadjuvant chemotherapy with high cross-cohort consistency (I²=0%), and generalises to the large I-SPY2 platform trial. The analytical pipeline is fully reproducible from publicly available data.

## Figures

```{r fig-study-design}
#| label: fig-study-design
#| fig-cap: "**Figure 1. Study overview.** Schematic of the Core-PAM derivation and validation pipeline, cohort roles, sample sizes, and frozen analytical parameters."
f <- file.path(ROOT, "figures/main/en/png/Fig1_StudyDesign_EN.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r fig-km-validation}
#| label: fig-km-validation
#| fig-cap: "**Figure 2. Kaplan-Meier survival curves — validation cohorts.** Patients dichotomised at the intra-cohort median Core-PAM score. Left: TCGA-BRCA (OS, N=1,072); Centre: METABRIC (DSS, N=1,978); Right: GSE20685 (OS, N=327). HR per 1 SD with 95%CI shown. Shading: 95%CI of KM estimate. Numbers at risk shown below x-axis."
for (f in c(file.path(ROOT, "figures/main/en/png/Fig2_KM_TCGA_BRCA_OS_EN.png"),
            file.path(ROOT, "figures/main/en/png/Fig2_KM_METABRIC_DSS_EN.png"),
            file.path(ROOT, "figures/main/en/png/Fig2_KM_GSE20685_OS_EN.png"))) {
  if (file.exists(f)) knitr::include_graphics(f)
}
```

```{r fig-km-training}
#| label: fig-km-training
#| fig-cap: "**Figure 3. Kaplan-Meier survival curve — SCAN-B training cohort.** OS dichotomised at the intra-cohort median Core-PAM score (N=3,069). Note: C-index in training is in-sample (0.698); out-of-fold C-index = 0.670."
f <- file.path(ROOT, "figures/main/en/png/Fig3_KM_SCANB_OS_CorePAM.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r fig-forest-meta}
#| label: fig-forest-meta
#| fig-cap: "**Figure 4. Forest plot — random-effects meta-analysis.** HR per 1 SD Core-PAM score across three validation cohorts plus pooled random-effects estimate. REML estimator."
f <- file.path(ROOT, "figures/main/en/png/Fig4_Meta_Forest_HR_per1SD_CorePAM.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r fig-incremental}
#| label: fig-incremental
#| fig-cap: "**Figure 5. Incremental value and calibration.** A: ΔC-index (Core-PAM added to CORE-A clinical model) with 95% bootstrap CI across cohorts. B: Calibration curves at 60 months (24m for TCGA-BRCA)."
for (f in c(file.path(ROOT, "figures/main/en/png/Fig5_DeltaCindex_COREA_vs_COREAplus_CorePAM.png"),
            file.path(ROOT, "figures/main/en/png/Fig5_Calibration_60m_Panels_CorePAM.png"))) {
  if (file.exists(f)) knitr::include_graphics(f)
}
```

```{r fig-pcr-forest}
#| label: fig-pcr-forest
#| fig-cap: "**Figure 6. Forest plot — Core-PAM and pCR.** OR per 1 SD Core-PAM score (logistic regression) in four neoadjuvant chemotherapy cohorts, with random-effects pooled estimate (DerSimonian-Laird; I²=0%)."
f <- file.path(ROOT, "figures/pcr/en/png/Fig_pCR1_Forest_OR_EN.png")
if (file.exists(f)) knitr::include_graphics(f)
```

\newpage

## Supplementary Figures

```{r figs1-pareto}
#| label: figs1-pareto
#| fig-cap: "**Figure S1. Pareto frontier: gene count vs. OOF C-index.** Each point represents a unique df level (number of non-zero genes) along the glmnet regularisation path. Dashed red line: non-inferiority boundary (C_max − 0.010). Selected Core-PAM point indicated."
f <- file.path(ROOT, "figures/supp/en/png/FigS_Pareto_CorePAM_EN.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r figs2-weights}
#| label: figs2-weights
#| fig-cap: "**Figure S2. Core-PAM gene weights (lollipop plot).** Elastic-net Cox regression coefficients. Positive weights (red): higher expression associated with worse survival. Negative weights (blue): higher expression associated with better survival. Genes ordered by absolute weight."
f <- file.path(ROOT, "figures/supp/en/png/FigS_Weights_CorePAM_Lollipop.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r figs3-correlation}
#| label: figs3-correlation
#| fig-cap: "**Figure S3. Off-diagonal Core-PAM score correlations between cohorts.** Spearman rho computed on paired percentiles across cohorts. Diagonal excluded (no self-autocorrelation). High correlation indicates cross-platform consistency of score distributions."
f <- file.path(ROOT, "figures/supp/en/png/FigS3_Correlation_OffDiagonal.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r figs5-metabric-sensitivity}
#| label: figs5-metabric-sensitivity
#| fig-cap: "**Figure S5. METABRIC sensitivity analyses.** A: OS endpoint (primary = DSS). B: Fine-Gray competing risks model. C: Quartile Kaplan-Meier (DSS)."
f <- file.path(ROOT, "figures/supp/en/png/FigS5_METABRIC_Sensitivity_EN.png")
if (file.exists(f)) knitr::include_graphics(f)
```

```{r figs6-tcga-24m}
#| label: figs6-tcga-24m
#| fig-cap: "**Figure S6. TCGA-BRCA 24-month sensitivity analysis.** Patients censored at 24 months. HR=1.64 (95%CI 1.24–2.16), p=0.00047. Mitigates the structural limitation of short follow-up in TCGA-BRCA."
f <- file.path(ROOT, "figures/supp/en/png/FigS6_TCGA_24m_Sensitivity_EN.png")
if (file.exists(f)) knitr::include_graphics(f)
```

\newpage

## Declarations

**Ethics approval and consent to participate:** Not applicable. This study uses exclusively publicly available de-identified data (GEO, GDC, cBioPortal). No new human subjects research was conducted.

**Consent for publication:** Not applicable.

**Competing interests:** The author declares no competing interests.

**Funding:** No external funding was received. This study was conducted independently by the author.

**Authors' contributions:** RB designed the study, performed all analyses, and wrote the manuscript.

**Acknowledgements:** The author thanks the SCAN-B, TCGA, METABRIC, and I-SPY consortia for making their data publicly available.

**Data availability:** All raw data are available through GEO (GSE96058, GSE25066, GSE20194, GSE32646, GSE22226, GSE194040), GDC (TCGA-BRCA), and cBioPortal (METABRIC). Analysis code and processed results are available at <https://github.com/RafaelBotan/Phd-Genomic/tree/main>.

## References

::: {#refs}
:::
